---
title: "Week 1 - Homework"
author: "STAT 420, Summer 2020, D. Unger"
date: 'Shubham Goel (sgoel11)'
output:
  html_document: 
    theme: readable
    toc: yes
    toc_depth: '2'  
  pdf_document: default
urlcolor: cyan
---

```{css, echo=FALSE}
.solution {
background-color: #e6ffe6;
}
```

```{r setup, include=TRUE, echo=FALSE}
  knitr::opts_chunk$set(cache=TRUE)
  knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_chunk$set(message = FALSE)
  knitr::opts_chunk$set(warning = FALSE)
  knitr::opts_chunk$set(fig.height = 6, fig.width = 8, fig.align = "center")
```
## Exercise 1 (Subsetting and Statistics)

For this exercise, we will use the `msleep` dataset from the `ggplot2` package.

**(a)** Install and load the `ggplot2` package. **Do not** include the installation command in your `.Rmd` file. (If you do it will install the package every time you knit your file.) **Do** include the command to load the package into your environment.

``` {r}
library(ggplot2)
library(dplyr)
library(MASS)
```

**(b)** Note that this dataset is technically a `tibble`, not a data frame. How many observations are in this dataset? How many variables? What are the observations in this dataset?

``` {r}
msleep
```

The dataset contains 83 observations with 11 variables. The observations are mammals and various characteristics about their sleep such as sleep_rem hours and total sleep_cycle hours. It also contains the total no. of hours the mammal stays awake, its body and brain weight.

**(c)** What is the mean hours of REM sleep of individuals in this dataset?

``` {r}
mean(msleep$sleep_rem, na.rm=TRUE)
```

1.875 hours is the mean REM sleep of individuals.

**(d)** What is the standard deviation of brain weight of individuals in this dataset?

``` {r}
sd(msleep$brainwt, na.rm = TRUE)
```

Brain weight's standard deviation is 0.976.

**(e)** Which observation (provide the `name`) in this dataset gets the most REM sleep?

```{r}
msleep[which.max(msleep$sleep_rem),]$name
```

**(f)** What is the average bodyweight of carnivores in this dataset?

```{r}
library(dplyr)
mean(filter(msleep, order == "Carnivora")$bodywt)
```

***

## Exercise 2 (Plotting)

For this exercise, we will use the `birthwt` dataset from the `MASS` package.

**(a)** Note that this dataset is a data frame and all of the variables are numeric. How many observations are in this dataset? How many variables? What are the observations in this dataset?
```{r}
nrow(birthwt)
ncol(birthwt)
```

There are 189 observations in this dataset consisting of 10 variables. The observations are observations made around pregnancy regarding the mother's age, smoking status, child's weight, etc. collected at Baystate Medical Center, Springfield, Mass during 1986.

**(b)** Create a scatter plot of birth weight (y-axis) vs mother's weight before pregnancy (x-axis). Use a non-default color for the points. (Also, be sure to give the plot a title and label the axes appropriately.) Based on the scatter plot, does there seem to be a relationship between the two variables? Briefly explain.

```{r}
plot(bwt ~ lwt, data = birthwt,
      ylab = "Birth weight in grams",
      xlab = "Mother's weight in pounds before pregnancy",
      main = "Birth weight in grams vs Mother's weight before pregnancy",
      pch  = 20,
      cex  = 1,
      col  = "dodgerblue")
```
Based on the graph, there seems to be a slightly positive relation between the birth weight and the mother's weight just before pregnancy. The correlation value would be quite **low**, as infered from the plot.


**(c)** Create a scatter plot of birth weight (y-axis) vs mother's age (x-axis). Use a non-default color for the points. (Also, be sure to give the plot a title and label the axes appropriately.) Based on the scatter plot, does there seem to be a relationship between the two variables? Briefly explain.
```{r}
plot(bwt ~ age, data = birthwt,
      xlab = "Mother's age in years",
      ylab = "Child's birth weight in grams",
      main = "Child's birth weight in grams vs Mother's age in years",
      pch  = 20,
      cex  = 1,
      col  = "dodgerblue")
```
Based on the graph, there seems to be no relation between the birth weight and the mother's age.


**(d)** Create side-by-side boxplots for birth weight grouped by smoking status. Use non-default colors for the plot. (Also, be sure to give the plot a title and label the axes appropriately.) Based on the boxplot, does there seem to be a difference in birth weight for mothers who smoked? Briefly explain.
```{r}
boxplot(bwt ~ smoke, data = birthwt,
     xlab   = "Smoking status (1 is for yes)",
     ylab   = "Birth weight in grams",
     main   = "Birth weight grouped by smoking status",
     pch    = 20,
     cex    = 2,
     col    = "darkorange",
     border = "dodgerblue")
```
The mean birth weight is higher in the case of women who didn't smoke during pregnancy, as well as the IQR range. The maximum weight for the non-smoking category is ~800gms higher than the smoking one, but the minimum is slightly lower.

***

## Exercise 3 (Importing Data, More Plotting)

For this exercise we will use the data stored in [`nutrition-2018.csv`](nutrition-2018.csv). It contains the nutritional values per serving size for a large variety of foods as calculated by the USDA in 2018. It is a cleaned version totaling 5956 observations and is current as of April 2018.

The variables in the dataset are:

- `ID` 
- `Desc` - short description of food
- `Water` - in grams
- `Calories` - in kcal
- `Protein` - in grams
- `Fat` - in grams
- `Carbs` - carbohydrates, in grams
- `Fiber` - in grams
- `Sugar` - in grams
- `Calcium` - in milligrams
- `Potassium` - in milligrams
- `Sodium` - in milligrams
- `VitaminC` - vitamin C, in milligrams
- `Chol` - cholesterol, in milligrams
- `Portion` - description of standard serving size used in analysis

```{r}
library(readr)
nutrition_2018 <- read_csv("nutrition-2018.csv")
nutrition_2018
```

**(a)** Create a histogram of `Calories`. Do not modify `R`'s default bin selection. Make the plot presentable. Describe the shape of the histogram. Do you notice anything unusual?
```{r}
hist(nutrition_2018$Calories,
     xlab   = "Calories",
     main   = "Histogram of Calories",
     col    = "dodgerblue",
     border = "darkorange")
```
The graph is largely uniform (and does not follow a normal distribution). Most of the foods contain less than 600 calories, with an extreme minority containing >800 calories. These seem to be foods such as oils, beef fats and chicken fats, consisting of very high quantities of fats.

**(b)** Create a scatter plot of calories (y-axis) vs protein (x-axis). Make the plot presentable. Do you notice any trends? Do you think that knowing only the protein content of a food, you could make a good prediction of the calories in the food?
```{r}
plot(Calories ~ Protein, data = nutrition_2018,
      xlab = "Protein",
      ylab = "Calories",
      main = "Calories vs Protein",
      pch  = 20,
      cex  = 1,
      col  = "dodgerblue")
```
There seems to be no correlation between the amount of protein present in a food and the calorie amount. However, as the protein amount increases, the spread of the calories in that food item seems to decreease. No, calorie prediction cannot be made based on the amount of protein (probably beause the amount of fats and carbs are not fixed, and they affect the total calories)

**(c)** Create a scatter plot of `Calories` (y-axis) vs `4 * Protein + 4 * Carbs + 9 * Fat` (x-axis). Make the plot presentable. You will either need to add a new variable to the data frame, or use the `I()` function in your formula in the call to `plot()`. If you are at all familiar with nutrition, you may realize that this formula calculates the calorie count based on the protein, carbohydrate, and fat values. You'd expect then that the result here is a straight line. Is it? If not, can you think of any reasons why it is not?

```{r}
nutrition_2018$estimated_calories = (4 * nutrition_2018$Protein) + (4 * nutrition_2018$Carbs) + (9 * nutrition_2018$Fat)
plot(Calories ~ estimated_calories, data = nutrition_2018,
      xlab = "4 * Protein + 4 * Carbs + 9 * Fat",
      ylab = "Calories",
      main = "Calories vs 4 * Protein + 4 * Carbs + 9 * Fat",
      pch  = 20,
      cex  = 1,
      col  = "dodgerblue")
```
The scatter plot is largely a straight line, which is expected as the formula is used to estimate calories present inside a food. The deviations are because of the fact that the coefficients 4,4 and 9 are roughly average estimates for proteins, carbs and fats and it is very much possible that a particular fat source, for example, might have 11 or 7 kcals per gram in it. 

***

## Exercise 4 (Writing and Using Functions)

For each of the following parts, use the following vectors:

```{r}
a = 1:10
b = 10:1
c = rep(1, times = 10)
d = 2 ^ (1:10)
```

**(a)** Write a function called `sum_of_squares`.

- Arguments:
    - A vector of numeric data `x`
- Output:
    - The sum of the squares of the elements of the vector $\sum_{i = 1}^n x_i^2$
    
Provide your function, as well as the result of running the following code:

```{r}
sum_of_squares = function(x){
  return(sum(x**2))
  
}
sum_of_squares(x = a)
sum_of_squares(x = c(c, d))
```

**(b)** Using only your function `sum_of_squares()`, `mean()`, `sqrt()`, and basic math operations such as `+` and `-`, calculate

\[
\sqrt{\frac{1}{n}\sum_{i = 1}^n (x_i - 0)^{2}}
\]

where the $x$ vector is `d`.

```{r}
sqrt(sum_of_squares(d-0)/length(d))
```


**(c)** Using only your function `sum_of_squares()`, `mean()`, `sqrt()`, and basic math operations such as `+` and `-`, calculate

\[
\sqrt{\frac{1}{n}\sum_{i = 1}^n (x_i - y_i)^{2}}
\]

where the $x$ vector is `a` and the $y$ vector is `b`.

```{r}
sqrt(sum_of_squares(a-b)/length(a))
```


***

## Exercise 5 (More Writing and Using Functions)

For each of the following parts, use the following vectors:

```{r}
set.seed(42)
x = 1:100
y = rnorm(1000)
z = runif(150, min = 0, max = 1)
```

**(a)** Write a function called `list_extreme_values`.

- Arguments:
    - A vector of numeric data `x`
    - A positive constant, `k`, with a default value of `2`
- Output:
    - A list with two elements:
        - `small`, a vector of elements of `x` that are $k$ sample standard deviations less than the sample mean. That is, the observations that are smaller than $\bar{x} - k \cdot s$.
        - `large`, a vector of elements of `x` that are $k$ sample standard deviations greater than the sample mean. That is, the observations that are larger than $\bar{x} + k \cdot s$.

Provide your function, as well as the result of running the following code:

```{r}
list_extreme_values = function(x, k = 2){
  mean = mean(x)
  sd = sd(x)
  small = x[x<(mean-k*sd)]
  big = x[x>(mean+k*sd)]
  final_list = list('small' = small, 'big' = big)
  return(final_list)
}
list_extreme_values(x = x, k = 1)
list_extreme_values(x = y, k = 3)
list_extreme_values(x = y, k = 2)
list_extreme_values(x = z, k = 1.5)
```

**(b)** Using only your function `list_extreme_values()`, `mean()`, and basic list operations, calculate the mean of observations that are greater than 1.5 standard deviation above the mean in the vector `y`.

```{r}
mean(list_extreme_values(x = y, k = 1.5)$big)
```

This is the mean. 

***

