---
title: "Week 2 - Homework"
author: "STAT 420, Summer 2020, D. Unger"
date: 'Shubham Goel (sgoel11)'
output:
  html_document: 
    theme: readable
    toc: yes
    toc_depth: '2'  
  pdf_document: default
urlcolor: cyan
---

```{css, echo=FALSE}
.solution {
background-color: #e6ffe6;
}
```

```{r setup, include=TRUE, echo=FALSE}
  knitr::opts_chunk$set(cache=TRUE)
  knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_chunk$set(message = FALSE)
  knitr::opts_chunk$set(warning = FALSE)
  knitr::opts_chunk$set(fig.height = 6, fig.width = 8, fig.align = "center")
```

## Exercise 1 (Using `lm`)

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

**(a)** Suppose we would like to understand the size of a cat's heart based on the body weight of a cat. Fit a simple linear model in `R` that accomplishes this task. Store the results in a variable called `cat_model`. Output the result of calling `summary()` on `cat_model`.

```{r}
library(MASS)
cat_model = lm(Hwt ~ Bwt, data = cats)
summary(cat_model)
```

**(b)** Output only the estimated regression coefficients. Interpret $\hat{\beta_0}$ and $\beta_1$ in the *context of the problem*. Be aware that only one of those is an estimate.
```{r}
cat_model$coefficients
```
The Bwt coefficient tells us that for a 1 unit increase in the bodyweight (in kg), the heart size grows by 4.03 units (in grams). The intercept is the constant value of heart weight which needs to be added to the Bwt term to get an actual estimate for the heart weight. It tells the value the heart weight would take when the body weight is 0 (not practical in this case).

**(c)** Use your model to predict the heart weight of a cat that weights **3.1** kg. Do you feel confident in this prediction? Briefly explain.
```{r}
predict(cat_model, newdata = data.frame(Bwt=3.1))
```
Since our original data has bodyweight in range from 2 to 3.9kgs, we are confident about this result.

**(d)** Use your model to predict the heart weight of a cat that weights **1.5** kg. Do you feel confident in this prediction? Briefly explain.
```{r}
predict(cat_model, newdata = data.frame(Bwt=1.5))
```
This value goes out of bounds from our model's input training data, so we're not very confident about the predictions made here.

**(e)** Create a scatterplot of the data and add the fitted regression line. Make sure your plot is well labeled and is somewhat visually appealing.
```{r}
plot(Hwt ~ Bwt, data = cats,
     xlab = "Body weight (in kg)",
     ylab = "Heart weight (in grams)",
     main = "Heart rate vs Bodyweight",
     pch  = 20,
     cex  = 2,
     col  = "grey")
abline(cat_model, lwd = 3, col = "darkorange")
```


**(f)** Report the value of $R^2$ for the model. Do so directly. Do not simply copy and paste the value from the full output in the console after running `summary()` in part **(a)**.
```{r}
summary(cat_model)$r.squared
```

***

## Exercise 2 (Writing Functions)

This exercise is a continuation of Exercise 1.

**(a)** Write a function called `get_sd_est` that calculates an estimate of $\sigma$ in one of two ways depending on input to the function. The function should take three arguments as input:

- `fitted_vals` - A vector of fitted values from a model
- `actual_vals` - A vector of the true values of the response
- `mle` - A logical (`TRUE` / `FALSE`) variable which defaults to `FALSE`

The function should return a single value:

- $s_e$ if `mle` is set to `FALSE`.
- $\hat{\sigma}$ if `mle` is set to `TRUE`.

```{r}
get_sd_est = function(fitted_vals, actual_vals, mle=FALSE){
  residuals = sum((fitted_vals-actual_vals)**2)
  #Returns s_e if mle is set to FALSE
  if(mle == FALSE){
    return(sqrt(residuals/(length(fitted_vals)-2)))
  }
  else{
    return(sqrt(residuals/length(fitted_vals)))
  }
}
```


**(b)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `FALSE`. Explain the resulting estimate in the context of the model.
```{r}
get_sd_est(cat_model$fitted.values, cats$Hwt)
```
This is the residual standard error, which shows the standard deviation of our residuals. A smaller error would indicate better predictions for the heart weight in this case. 

**(c)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `TRUE`. Explain the resulting estimate in the context of the model. Note that we are trying to estimate the same parameter as in part **(b)**.
```{r}
get_sd_est(cat_model$fitted.values, cats$Hwt, mle = TRUE)
```
This is the same residual standard error, but since we've divided the sum of residuals squared by a bigger number (without subtracting the degrees of freedoms), the error is smaller. This is a biased estimator of the standard error, which is why we prefer the **(b)** one, which is unbiased.

**(d)** To check your work, output `summary(cat_model)$sigma`. It should match at least one of **(b)** or **(c)**.
```{r}
summary(cat_model)$sigma
```
We see that it matches our output in **(b)**, which verifies our function.
***

## Exercise 3 (Simulating SLR)

Consider the model

\[
Y_i = 5 + -3 x_i + \epsilon_i
\]

with 

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 10.24)
\]

where $\beta_0 = 5$ and $\beta_1 = -3$.

This exercise relies heavily on generating random observations. To make this reproducible we will set a seed for the randomization. Alter the following code to make `birthday` store your birthday in the format: `yyyymmdd`. For example, [William Gosset](https://en.wikipedia.org/wiki/William_Sealy_Gosset), better known as *Student*, was born on June 13, 1876, so he would use:

```{r}
birthday = 19940401
set.seed(birthday)
```

**(a)** Use `R` to simulate `n = 25` observations from the above model. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
x = runif(n = 25, 0, 10)
```

You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Store the data frame this function returns in a variable of your choice. Note that this function calls $y$ `response` and $x$ `predictor`.
```{r}
sim_slr = function(x, beta_0 = 5, beta_1 = -3, sigma = 1) {
  n = length(x)
  epsilon = rnorm(n, mean = 0, sd = sigma)
  y = beta_0 + beta_1 * x + epsilon
  data.frame(predictor = x, response = y)
}
```

**(b)** Fit a model to your simulated data. Report the estimated coefficients. Are they close to what you would expect? Briefly explain.
```{r}
sim_data = sim_slr(x, sigma = sqrt(10.24))
sim_fit = lm(response ~ predictor, data = sim_data)
coef(sim_fit)
```
The estimated $\beta_0$ coefficient is 6.110 which is close to the actual $\beta_0$. Similarly, the $\beta_1$ estimate is -3.272 which is fairly close to the actual $\beta_1$. 

**(c)** Plot the data you simulated in part **(a)**. Add the regression line from part **(b)** as well as the line for the true model. Hint: Keep all plotting commands in the same chunk.
```{r}
beta_0 = 5
beta_1 = -3
plot(response ~ predictor, data = sim_data,
     xlab = "Simulated Predictor Variable",
     ylab = "Simulated Response Variable",
     main = "Simulated Regression Data",
     pch  = 20,
     cex  = 2,
     col  = "grey")
abline(sim_fit, lwd = 3, lty = 1, col = "darkorange")
abline(beta_0, beta_1, lwd = 3, lty = 2, col = "dodgerblue")
legend("topright", c("Estimate", "Truth"), lty = c(1, 2), lwd = 2,
       col = c("darkorange", "dodgerblue"))
```


**(d)** Use `R` to repeat the process of simulating `n = 25` observations from the above model $1500$ times. Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. Some hints:

- Consider a `for` loop.
- Create `beta_hat_1` before writing the `for` loop. Make it a vector of length $1500$ where each element is `0`.
- Inside the body of the `for` loop, simulate new $y$ data each time. Use a variable to temporarily store this data together with the known $x$ data as a data frame.
- After simulating the data, use `lm()` to fit a regression. Use a variable to temporarily store this output.
- Use the `coef()` function and `[]` to extract the correct estimated coefficient.
- Use `beta_hat_1[i]` to store in elements of `beta_hat_1`.
- See the notes on [Distribution of a Sample Mean](http://daviddalpiaz.github.io/appliedstats/introduction-to-r.html#distribution-of-a-sample-mean) for some inspiration.

You can do this differently if you like. Use of these hints is not required.

```{r}
beta_hat1 = rep(0,1500)
for (i in 1:1500){
  sim_data = sim_slr(x, sigma = sqrt(10.24))
  sim_fit = lm(response ~ predictor, data = sim_data)
  beta_hat1[i] = coef(sim_fit)[2]
}
```

**(e)** Report the mean and standard deviation of `beta_hat_1`. Do either of these look familiar?
```{r}
mean(beta_hat1)
sd(beta_hat1)
```
The mean is extemely close to the actual coefficient for `beta_hat_1`. 

**(f)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.
```{r}
hist(beta_hat1,
     xlab   = "beta_hat1",
     main   = "Histogram of beta_hat1",
     col    = "dodgerblue",
     border = "darkorange")
```
The graph resembles a normal distribution, which is expected as the distribution of a sample mean from any distribution, is a normal distribution.

***

## Exercise 4 (Be a Skeptic)

Consider the model

\[
Y_i = 3 + 0 \cdot x_i + \epsilon_i
\]

with

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 4)
\]

where $\beta_0 = 3$ and $\beta_1 = 0$.

Before answering the following parts, set a seed value equal to **your** birthday, as was done in the previous exercise.

```{r}
birthday = 19940401
set.seed(birthday)
```

**(a)** Use `R` to repeat the process of simulating `n = 75` observations from the above model $2500$ times. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
x = runif(n = 75, 0, 10)
```

Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Hint: Yes $\beta_1 = 0$.

```{r}
beta_hat_1 = rep(0,1500)
for (i in 1:1500){
  sim_data = sim_slr(x, beta_0 = 3, beta_1 = 0, sigma = 2)
  sim_fit = lm(response ~ predictor, data = sim_data)
  beta_hat_1[i] = coef(sim_fit)[2]
}
```

**(b)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.

```{r}
hist(beta_hat_1,
     xlab   = "beta_hat1",
     main   = "Histogram of beta_hat1",
     col    = "dodgerblue",
     border = "darkorange")
```
The shape again resembles a normal distribution. A distribution of sample means taken from any distribution is a normal distribution.

**(c)** Import the data in [`skeptic.csv`](skeptic.csv) and fit a SLR model. The variable names in `skeptic.csv` follow the same convention as those returned by `sim_slr()`. Extract the fitted coefficient for $\beta_1$.
```{r}
library(readr)
skeptic <- read_csv("skeptic.csv")
skeptic_fit = lm(response ~ predictor, data = skeptic)
coef(skeptic_fit)[2]
```


**(d)** Re-plot the histogram from **(b)**. Now add a vertical red line at the value of $\hat{\beta_1}$ in part **(c)**. To do so, you'll need to use `abline(v = c, col = "red")` where `c` is your value.
```{r}
skeptic_beta_hat_1 = coef(skeptic_fit)[2]
hist(beta_hat_1,
     xlab   = "beta_hat1",
     main   = "Histogram of beta_hat1",
     col    = "dodgerblue",
     border = "darkorange")
abline(v = coef(skeptic_fit)[2], col = "red")
```


**(e)** Your value of $\hat{\beta_1}$ in **(c)** should be negative. What proportion of the `beta_hat_1` values is smaller than your $\hat{\beta_1}$? Return this proportion, as well as this proportion multiplied by `2`.
```{r}
sum(beta_hat_1<skeptic_beta_hat_1)/length(beta_hat_1)
sum(beta_hat_1<skeptic_beta_hat_1)/length(beta_hat_1)*2
```


**(f)** Based on your histogram and part **(e)**, do you think the [`skeptic.csv`](skeptic.csv) data could have been generated by the model given above? Briefly explain.

It looks **extremely unlikely** that the data was generated by the model, as the `beta_hat_1` value obtained obtained from fitting on the skeptic data doesn't fall in the centre of the distribution of the actual `beta_hat_1` means distribution. 

***

## Exercise 5 (Comparing Models)

For this exercise we will use the `Ozone` dataset from the `mlbench` package. You should use `?Ozone` to learn about the background of this dataset. You may need to install the `mlbench` package. If you do so, do not include code to install the package in your `R` Markdown document.

```{r}
library(mlbench)
```

For simplicity, we will perform some data cleaning before proceeding.

```{r}
data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

We have:

- Loaded the data from the package
- Subset the data to relevant variables
    - This is not really necessary (or perhaps a good idea) but it makes the next step easier
- Given variables useful names
- Removed any observation with missing values
    - This should be given much more thought in practice

For this exercise we will define the "Root Mean Square Error" of a model as

\[
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2}.
\]

**(a)** Fit three SLR models, each with "ozone" as the response. For the predictor, use "wind speed," "humidity percentage," and "temperature" respectively. For each, calculate $\text{RMSE}$ and $R^2$. Arrange the results in a markdown table, with a row for each model. Suggestion: Create a data frame that stores the results, then investigate the `kable()` function from the `knitr` package.

```{r}
library(Metrics)
library(knitr)
wind_model = lm(ozone ~ wind, data = Ozone)
humidity_model = lm(ozone ~ humidity, data = Ozone)
temp_model = lm(ozone ~ temp, data = Ozone)
rmse_vals = c(rmse(Ozone$ozone, wind_model$fitted.values), rmse(Ozone$ozone, humidity_model$fitted.values), rmse(Ozone$ozone, temp_model$fitted.values))
r_squared_vals = c(summary(wind_model)$r.squared, summary(humidity_model)$r.squared, summary(temp_model)$r.squared)
df = data.frame(rmse_vals, r_squared_vals, row.names = c("Wind model", "Humidity model", "Temperature model"))
kable(df)
```


**(b)** Based on the results, which of the three predictors used is most helpful for predicting ozone readings? Briefly explain.

The temperature variable has the lowest `RMSE` and the highest `R^2` value, which makes it the most useful for predicting the ozone readings. The model built using this variable is able to explain the highest variance in the data. The wind model, inversely, is the worst performing one.

***
